Multiple imputation of incomplete multilevel data using Heckman selection models

### 2022-10-24 1.Paper/Paper.Rmd
update pdf

Registries are notoriously prone to incomplete variables that are MNAR, due to the complex recording process <!--[@liu2021]-->, e.g. laboratory tests are taken only in certain patients based on sign and symptom information that is often incomplete or not recorded. Also IPD may suffer from MNAR, for example when study participants who experience unfavorable results drop out of the study or also due to heterogeneity of the primary objective or resources of the studies involved, which may result in variables relevant to explain the missing process not being recorded.

@@

In clustered data sets, multilevel imputation methods are needed to properly propagate uncertainty within and across clusters [audigier_etal18]. However, to our knowledge, existing multilevel imputation methods mainly focus  on situations where data are MAR, and do not adopt Heckman selection models. Although Hammon and Zinn (2020) [@hammon_zinn20] recently proposed an extension that allows for the inclusion of random intercept effects, it can only be used for binary missing variables and assume that the effect of explanatory variables on the missingness mechanisms is common across clusters.

@@

# Simulation study
#### Aim
We design a simulation study aimed to compare the performance of imputation methods when they impute a missing variable that comes from a hierarchical dataset and follows a MNAR mechanism.

@@

@@

#### Results
```{r setup, include=FALSE} 
knitr::opts_chunk$set(warning = FALSE, message = FALSE) 
```

# ```{r, include=FALSE}
# library(here)
# source(here("2.Simulation", "simplot.R"),local =knitr::knit_global())
# ```

@@

##### Results M(N)AR scenarios

# ```{r rhocont,fig.pos="h",fig.cap="\\label{fig:rhocont} Comparison of methods for continuous incomplete variable under systematical missingness by varing $\rho$, the dashed line depicts the true simulated parameter value",echo=FALSE,fig.keep="last"}
# plot_cont_rho
# ```

Figure \ref{fig:rhocont} shows the results of simulations where the missing variable was continuous.In the MAR scenario, i.e. when $\rho=0$, all imputation methods provide similar unbiased estimates of the coefficient parameters, but as rho increases, i.e. the mechanism becomes MNAR, the estimates for the complete case analysis and the MAR-IPD imputation method turns more biased.
Overall, both Heckman-based imputation (1l. and 2l.) models provide less biased estimates of the coefficient parameters in MNAR scenarios, but the random effects estimates on the 1l.heckman are far away from the true values as no cluster information was considered at all. The precision of the Heckman IPD separated seems it is better than the Heckman IPD full version, specially in the MAR scenarios, when we checked simulation results it seems that the high variability on the full version is related to simulations in which singularity fit that it is an indication that the model is overfitted. Regarding the coverage, the Heckman-IPD imputation method provides the best coverage values across all the $\rho$ scenarios with a coverage level up to 95%.

The Heckman-IPD method, by incorporating individual and marginal level information, allows a bias-variance trade off, thus, as it was observed in Table \ref{tab:rhotable} the Heckman IPD separated resulted in better estimates in terms of RMSE than the other methods evaluated.
In particular, the method provides an advantage over the Heckman-single method in systematic missingness scenarios, since the latter does not allow to specify any cluster information implicitly, i.e. by adding a group variable in the imputation model. This can be seen in the estimates of the random effects parameters of the Heckman-single model, which are more biassed than those of the other methods in which cluster information was included.
Moreover, the method could be affected by the ratio of the pooled data (See Annex Figure\ref{fig:continuous3}). To demonstrate this, we increased the variance of the random effects in the simulated data, reducing the intra-cluster correlation (ICC) from 0.33 to 0.23. This resulted in less biassed estimates for both coefficient and random effects parameters, along with narrower confidence intervals.

##### Results sensitivity to number of clusters and sample size of clusters
We evaluated how robust our method was to variations in the sample size of the studies included in the IPD and also the number of clusters.
Our method allows for the inclusion of studies with different sample sizes, but to facilitate the evaluation of the method, in this simulation study we used 10 studies each with a fixed sample size of 1000 observations.

<!-- ```{r rhosize,fig.pos="h",fig.cap="\\label{fig:size} Comparison of methods for continuous incomplete variable under systematical missingness by varing (number of studies- sample size per study), the dashed line depicts the true simulated parameter value",echo=FALSE,fig.keep="last"} -->
<!-- figure_final_s -->
<!-- ``` -->

When we reduced the sample size to 50 patients per study (left-most panel 10-50, Figure \ref{fig:rhosize}), we observed that the bias could be affected by some of the coefficient parameters, especially in MNAR scenarios. All simulated data were successfully imputed with the Heckman IPD methods but we observed that the precision of the parameters was drastically affected especially for the Heckman IPD full method, due to singularity fitting problems in the estimation of the mixed model, so we prefer not to show it in the graph.

On the other hand, by increasing the number of studies to 100 (rightmost panel 100-1000, Figure \ref{fig:rhosize}) we observe a higher precision of the estimates. There was an increase in the coverage of the estimates in the MAR scenario, i.e. rho=0, as the number of studies increases, which is in line with Audigier's results. However, in the case of MNAR, i.e. rho=0.6, we unfortunately observed a decrease in coverage, which could be explained by an increased complexity of the Heckman model.

Furthermore, we observed that for the full Heckman IPD method, the number of successfully imputed datasets was only 90% of the simulated datasets. This could be explained by the fact that increasing the number of studies included in the analysis might make it more difficult to estimate the marginal parameters when applying the meta-analysis model with random effects, especially when considering maximum likelihood or restrictive maximum likelihood as the estimation method. To ensure the estimation of the marginal parameters, our method evaluates the marginal parameters using the method of moments when it is not possible to estimate them with other methods chosen by the user.  We use the method of moments estimation as an alternative estimation method, as it was previously evaluated by Audigier, and was found to provide robust results when estimating the marginal parameters.

@@

However for school children, prevalences estimated with the Heckman method were found to be higher in the Mid-East and Southwest regions than those obtained with the other methods, whereas in the East-Central region the estimates with the Heckman method are lower. Possible reasons for selection bias in surveys of this type have been suggested, for example, that daytime visits might favor measurement in sick school children who stay home, leading to overestimated prevalence results as found in the East-Central region. Nevertheless, we were unable to find information that suggests or confirms the direction in which malaria prevalence is driven by selection bias in this Uganda study or in other studies similar to this one.

>>>>

Registries are notoriously prone to incomplete variables that are MNAR, due to the complex recording process [@liu2021], e.g. laboratory tests are taken only in certain patients based on sign and symptom information that is often incomplete or not recorded. Also IPD may suffer from MNAR, for example when study participants who experience unfavorable results drop out of the study or also due to heterogeneity of the primary objective or resources of the studies involved, which may result in variables relevant to explain the missing process not being recorded.

@@

In clustered data sets, multilevel imputation methods are needed to properly propagate uncertainty within and across clusters [@audigier_etal18]. However, to our knowledge, existing multilevel imputation methods mainly focus  on situations where data are MAR, and do not adopt Heckman selection models. Although Hammon and Zinn (2020) [@hammon_zinn20] recently proposed an extension that allows for the inclusion of random intercept effects, it can only be used for binary missing variables and assume that the effect of explanatory variables on the missingness mechanisms is common across clusters.

@@

# Simulation study
\label{simulation}

#### Aim
We design a simulation study aimed to compare the performance of imputation methods when they impute a missing variable that comes from a hierarchical dataset and follows a MNAR mechanism.

@@

For coverage estimation, we obtain the confidence intervals from the Wald method. For the random effect variance, even when it is possible to estimate the confidence intervals through the profile or bootstrap method, we did not estimated them for computational reasons.

@@

#### Results

```{r setup, include=FALSE} 
knitr::opts_chunk$set(warning = FALSE, message = FALSE) 
```

<!-- ```{r, include=FALSE} -->
<!-- library(here) -->
<!-- source(here("2.Simulation", "simplot.R"),local =knitr::knit_global()) -->
<!-- ``` -->

@@

<!-- ##### Results M(N)AR scenarios -->

<!-- # ```{r rhocont,fig.pos="h",fig.cap="\\label{fig:rhocont} Comparison of methods for continuous incomplete variable under systematical missingness by varing $\rho$, the dashed line depicts the true simulated parameter value",echo=FALSE,fig.keep="last"} -->
<!-- # plot_cont_rho -->
<!-- # ``` -->

<!-- Figure \ref{fig:rhocont} shows the results of simulations where the missing variable was continuous.In the MAR scenario, i.e. when $\rho=0$, all imputation methods provide similar unbiased estimates of the coefficient parameters, but as rho increases, i.e. the mechanism becomes MNAR, the estimates for the complete case analysis and the MAR-IPD imputation method turns more biased. -->
<!-- Overall, both Heckman-based imputation (1l. and 2l.) models provide less biased estimates of the coefficient parameters in MNAR scenarios, but the random effects estimates on the 1l.heckman are far away from the true values as no cluster information was considered at all. The precision of the Heckman IPD separated seems it is better than the Heckman IPD full version, specially in the MAR scenarios, when we checked simulation results it seems that the high variability on the full version is related to simulations in which singularity fit that it is an indication that the model is overfitted. Regarding the coverage, the Heckman-IPD imputation method provides the best coverage values across all the $\rho$ scenarios with a coverage level up to 95%. -->

<!-- The Heckman-IPD method, by incorporating individual and marginal level information, allows a bias-variance trade off, thus, as it was observed in Table \ref{tab:rhotable} the Heckman IPD separated resulted in better estimates in terms of RMSE than the other methods evaluated. -->
<!-- In particular, the method provides an advantage over the Heckman-single method in systematic missingness scenarios, since the latter does not allow to specify any cluster information implicitly, i.e. by adding a group variable in the imputation model. This can be seen in the estimates of the random effects parameters of the Heckman-single model, which are more biassed than those of the other methods in which cluster information was included. -->
<!-- Moreover, the method could be affected by the ratio of the pooled data (See Annex Figure\ref{fig:continuous3}). To demonstrate this, we increased the variance of the random effects in the simulated data, reducing the intra-cluster correlation (ICC) from 0.33 to 0.23. This resulted in less biassed estimates for both coefficient and random effects parameters, along with narrower confidence intervals. -->

<!-- ##### Results sensitivity to number of clusters and sample size of clusters -->
<!-- We evaluated how robust our method was to variations in the sample size of the studies included in the IPD and also the number of clusters. -->
<!-- Our method allows for the inclusion of studies with different sample sizes, but to facilitate the evaluation of the method, in this simulation study we used 10 studies each with a fixed sample size of 1000 observations. -->

<!-- <!-- ```{r rhosize,fig.pos="h",fig.cap="\\label{fig:size} Comparison of methods for continuous incomplete variable under systematical missingness by varing (number of studies- sample size per study), the dashed line depicts the true simulated parameter value",echo=FALSE,fig.keep="last"} --> -->
<!-- <!-- figure_final_s --> -->
<!-- <!-- ``` --> -->

<!-- When we reduced the sample size to 50 patients per study (left-most panel 10-50, Figure \ref{fig:rhosize}), we observed that the bias could be affected by some of the coefficient parameters, especially in MNAR scenarios. All simulated data were successfully imputed with the Heckman IPD methods but we observed that the precision of the parameters was drastically affected especially for the Heckman IPD full method, due to singularity fitting problems in the estimation of the mixed model, so we prefer not to show it in the graph. -->

<!-- On the other hand, by increasing the number of studies to 100 (rightmost panel 100-1000, Figure \ref{fig:rhosize}) we observe a higher precision of the estimates. There was an increase in the coverage of the estimates in the MAR scenario, i.e. rho=0, as the number of studies increases, which is in line with Audigier's results. However, in the case of MNAR, i.e. rho=0.6, we unfortunately observed a decrease in coverage, which could be explained by an increased complexity of the Heckman model. -->

<!-- Furthermore, we observed that for the full Heckman IPD method, the number of successfully imputed datasets was only 90% of the simulated datasets. This could be explained by the fact that increasing the number of studies included in the analysis might make it more difficult to estimate the marginal parameters when applying the meta-analysis model with random effects, especially when considering maximum likelihood or restrictive maximum likelihood as the estimation method. To ensure the estimation of the marginal parameters, our method evaluates the marginal parameters using the method of moments when it is not possible to estimate them with other methods chosen by the user.  We use the method of moments estimation as an alternative estimation method, as it was previously evaluated by Audigier, and was found to provide robust results when estimating the marginal parameters. -->

@@

However for school children, prevalences estimated with the Heckman method were found to be higher in the Mid-East and Southwest regions than those obtained with the other methods, whereas in the East-Central region the estimates with the Heckman method are lower. Possible reasons for selection bias in surveys of this type have been suggested [@thedhsprogram2020], for example, that daytime visits might favor measurement in sick school children who stay home, leading to overestimated prevalence results as found in the East-Central region. Nevertheless, we were unable to find information that suggests or confirms the direction in which malaria prevalence is driven by selection bias in this Uganda study or in other studies similar to this one.
